---
title: "Introduction to SoRE: Annotation-Ranked Bayesian Fine-Mapping"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to SoRE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, fig.height = 4.5
)
```

## What SoRE does

SoRE (Sum of Ranked Effects) is a Bayesian fine-mapping prior that uses
functional annotation to localize causal variants within GWAS-identified
regions. Two design choices distinguish it from existing
annotation-informed methods:

1. **Ordering, not weights.** The prior is placed on the variant ordering
   through a Plackett-Luce distribution, so it is invariant to monotone
   rescaling of the annotation scores. There is no need to put
   heterogeneous annotation sources on a common scale.

2. **Locus-adaptive trust.** A single parameter $\tau^2$, estimated from
   the agreement between annotation ranks and current PIPs, controls
   how much the prior trusts the annotation at each locus. When
   annotation and data disagree, $\tau^2$ grows and the output reverts
   to SuSiE-RSS.

The package exposes two interfaces:

- `vsore()` runs the empirical-Bayes variational implementation (V-SoRE).
  This is the recommended interface for genome-wide pipelines.
- `sore_gibbs()` runs the full blocked Gibbs sampler for the complete
  hierarchical posterior. Use this when you need full uncertainty
  quantification.

## A simple synthetic example

```{r}
library(SoRE)

set.seed(1)
p <- 200
n <- 50000
s0 <- 2

# AR(1) LD matrix
R <- 0.9 ^ abs(outer(1:p, 1:p, "-"))

# Two causal variants at positions 50 and 150
causal <- c(50, 150)
b <- numeric(p); b[causal] <- 5

# Draw z-scores under the RSS model
z <- as.numeric(R %*% b + crossprod(chol(R), rnorm(p)))

# SuSiE-RSS baseline (uniform weights)
fit_uniform <- vsore(z, R, anchor_weights = rep(1, p), n = n)
cat("Uniform weights: ", sum(fit_uniform$pip > 0.5), "variants with PIP > 0.5\n")

# Anchor weights that mark the causal variants as plausible
omega <- rep(1, p)
omega[causal] <- 100

fit_annot <- vsore(z, R, anchor_weights = omega, n = n)
cat("Annotated:       ", sum(fit_annot$pip > 0.5), "variants with PIP > 0.5\n")
cat("tau2 =", round(fit_annot$tau2, 3), "\n")
```

## Credible sets

```{r}
cs_uniform <- credible_sets(fit_uniform$pip, coverage = 0.95, R = R,
                            min_purity = 0.5)
cs_annot   <- credible_sets(fit_annot$pip,   coverage = 0.95, R = R,
                            min_purity = 0.5)
cs_uniform$size
cs_annot$size
```

## Interpreting the tau-squared diagnostic

```{r}
tau2_diagnostic(c(0.1, 0.5, 1.3, NA))
```

A `tau2` below 0.2 indicates strong annotation-data concordance, values in
[0.2, 1] indicate moderate concordance, and values above 1 indicate that
the data have substantially overridden the annotation ordering. A real
locus with tau2 above 1 should be treated as effectively unannotated.

## Building anchor weights from multiple sources

```{r}
set.seed(1)
p <- 200
omega <- build_anchor_weights(
  coding     = sample(1:3, p, replace = TRUE, prob = c(0.95, 0.04, 0.01)),
  chrom_open = runif(p) < 0.05,
  conserved  = runif(p) < 0.10,
  z_prior    = rnorm(p, sd = 0.5)
)
summary(omega)
```

## Citation

If you use SoRE in your work, please cite:

Chen, S. (2026). Bayesian Fine-Mapping with an Annotation Ranking Prior
Using GWAS Summary Statistics. *Biometrics*.
